HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-25 docs/HLL_exp_bad/main.tex
tweaks

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Durand and Flajolet, 2003) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm, an adversary who has some flexibility in choosing what inputs will be recorded can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary can make HLL underestimate true cardinality by a factor this is exponential in the number of buckets used by the algorithm.

@@

The work of~\cite{cardestprivacy} breaks the privacy of HLL and concludes that having the ability to test the insertion \kenny{what does this mean? see other comment later} of a targeted item gives information about whether the item was originally present in the input data set. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality.

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of losing mergability) or in addition to an unsalted sketch (inducing a memory overhead). 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

\subsection{Adversarial Setting}
We model three distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, S3 makes the strongest assumptions about the adversarial capabilities, since it combines the knowledge of adversarys from S1 and S2.

\begin{itemize}
\item[S1:] the adversary has black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item; 
\item[S2:] the adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use.
\item[S3:] the adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate.
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. We do not put our focus on S1 as this previous work in~\cite{hllvuln} fully exploits this scenario. Their attack works as follows: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained''. It is unclear what this means in practice, since after insertion into the HLL, an item cannot usually be removed by the adversary. \kenny{Mathilde: can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more. Maybe it also applies to the ``privacy'' paper.}

We start by presenting our attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln}, and then use the additional information the adversary has access to in scenario S3 to make the attack even more powerful.

@@

\noindent\paragraph{Scenario S3:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the count values per bucket in the HLL sketch). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

@@

In scenario S3, we compute the average work required of the adversary to find an input $x$ such that $h(x)$ meets the attack's requirements. Let us assume for the moment that $c_i \ge 1$ for all buckets $i$. Assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increase the estimator is the probability that the  rightmost $\ell$ bits of $h(x)$ have $c_i$ or less leading zeroes; this probability is equal to $1-\frac{1}{2^{c_i+1}}$ assuming $h$ behaves like a random function. Averaging over all $2^n$ buckets (and using the fact that the bucket choice is uniformly random since the bits of $h$ are independent and uniformly random), the probability that input $x$ does not increase the estimator is given by:
\begin{eqnarray*}
\sum_{i=1}^{2^n}\frac{1}{2^n} \cdot (1-\frac{1}{2^{c_i+1}}) & = & 1-\frac{1}{2}\cdot (\frac{1}{2^n}\cdot \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) \\
& = &1-(2H_c)^{-1} \\
\end{eqnarray*}
where $H_c$ is the harmonic mean of the counts $2^{c_1}, 2^{c_2}, \ldots, 2^{c_{2^n}}$.

@@

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. Unfortunately, the adversary in scenario S2 has no way to detect that he is attacking an empty HLL and cannot adapt his strategy, so all buckets hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S3 is able to detect that all buckets are empty and adapt his attack accordingly. The adversary can target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality estimate trade-offs are of course possible.

@@

We challenge the adversary to pick the largest possible number of items to add to the sketch while trying to keep the cardinality estimate as low as possible, either in the S2 or the S3 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.

Those empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the adversary is able to hide approximately half of the items' set under S2, and around 83\%%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$).
$\approx (1-\frac{1}{2\times H_c})$ with $H_c\approx\frac{1'000}{256}=3.9$, under S3.
Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by the Eq. (29) of \cite{hll2} or 1 in respectively S2 and S3.

@@

\begin{table}[h]
\caption{Result of the attacks over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S3} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'000 & 0 & 1'005 & 0 \\ \hline
    \textbf{\# Items added} & 3'655'744 & 3'655'740 & 6'090'003 & 14'220 \\ \hline
    \textbf{Final Est. Card.} & 1'011 & 367 & 1'005 & 1 \\ \hline
\end{tabular}
\label{table:tab1}
\end{table}

\begin{table}[h]
\caption{Result of the attacks in the setting of \cite{hllvuln} over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S3} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'004 & 0 & 998 & 0 \\ \hline
    \textbf{\# Items added} & 124'940 & 125'085 & 204'669 & 501 \\ \hline
    \textbf{Final Est. Card.} & 1'058 & 367 & 998 & 1 \\ \hline
\end{tabular}
\label{table:tab2}
\end{table}

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm, an adversary who has some flexibility in choosing what inputs will be recorded can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary can make HLL underestimate true cardinality by a factor that is exponential in the number of buckets used by the algorithm.

@@

The work of~\cite{cardestprivacy} studies the privacy properties of HLL in two different attack scenarios, referred to as \emph{insider} and \emph{external}. The attack target of~\cite{cardestprivacy} is different from ours, being concerned with loss of privacy, whereas we care about cardinality estimation. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm. In Section~\ref{sec:attacks} we will discuss the two attack scenarios from~\cite{cardestprivacy} in more detail. In particular, we will show that an attacker who can insert items into the HLL and who can access the HLL cardinality estimate (via typical APIs in HLL implementations) as permitted in the external scenario can actually realise the same attacks as the insider attacker can, at least insofar as cardinality estimate manipulation is concerned.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. They use an attack model that we consider to be quite artificial, since it gives the attacker the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. See further discussion in Section~\ref{sec:attacks}.

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of losing mergeability) or in addition to an unsalted sketch (inducing a memory overhead). 
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

\subsection{Adversarial Setting}
We model three distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, S4 makes the strongest assumptions about the adversarial capabilities, since it combines the knowledge of adversaries from S1 and S2. \kenny{Not really! Because S1 is actually quite unrealistic, as we will argue below.}

\begin{itemize}
\item[S1:] The adversary has black-box access to the HLL in use -- that is, the adversary can ask for the HLL's cardinality estimate at any point in time, and so is able to check the effect on the cardinality estimate when inserting an item. \kenny{TODO: clarify more what this is terms of a ''test oracle''; its stronger than described because the adverary can test the effect of an insertion without making the insertion!s}
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack.
\item[S3:] \kenny{TODO: The external attacker from~\cite{cardestprivacy}  described in terms of oracles.}
\item[S4:] the adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch. Recall that these buckets hold the intermediate estimators that are averaged in a later step to compute the final cardinality estimate. Again, an S4 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack, but it can compute these values for itself (assuming its attack proceeds without input from other users being added to the HLL sketch in parallel with the attack).
\end{itemize}

Scenarios S1 and S2 are presented as M2 and M1, respectively, in~\cite{hllvuln}. We do not put our focus on S1 as this previous work in~\cite{hllvuln} fully exploits this scenario. Their attack works as follows: ``after each insertion the cardinality is checked and the items which do not increment the HLL estimate are retained''. It is unclear what this means in practice, since after insertion into the HLL, an item cannot usually be removed by the adversary. \kenny{Mathilde: can you think about this sentence that I added? Do we need to rewrite S1? The scenario does not really make sense at all in practice. I think we might use this to beat up the other paper a bit more.}

\kenny{TODO: Show that S3 implies S4 $O(2^{n + \ell})$ precomputation and $O(\ell \cdot 2^{n})$ insert and read queries. So we. focus from now on S2 and S4 -- because S1 is unrealistic and S3 implies S4.}

We start by presenting our attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln}, and then use the additional information the adversary has access to in scenario S4 to make the attack even more powerful.

@@

\noindent\paragraph{Scenario S4:} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the count values per bucket in the HLL sketch). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

@@

In scenario S4, we compute the average work required of the adversary to find an input $x$ such that $h(x)$ meets the attack's requirements. Let us assume for the moment that $c_i \ge 1$ for all buckets $i$. Assume $h(x)$ is mapped to bucket $i$. The probability that $h(x)$ does not increase the estimator is the probability that the  rightmost $\ell$ bits of $h(x)$ have $c_i$ or less leading zeroes; this probability is equal to $1-\frac{1}{2^{c_i+1}}$ assuming $h$ behaves like a random function. Averaging over all $2^n$ buckets (and using the fact that the bucket choice is uniformly random since the bits of $h$ are independent and uniformly random), the probability that input $x$ does not increase the estimator is given by:
\begin{eqnarray*}
\sum_{i=1}^{2^n}\frac{1}{2^n} \cdot (1-\frac{1}{2^{c_i+1}}) & = & 1-\frac{1}{2}\cdot (\frac{1}{2^n}\cdot \sum_{i=1}^{2^n} {(2^{c_i})}^{-1}) \\
& = &1-(2H_c)^{-1} \\
\end{eqnarray*}
where $H_c$ is the harmonic mean of the counts $2^{c_1}, 2^{c_2}, \ldots, 2^{c_{2^n}}$.

@@

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. Unfortunately, the adversary in scenario S2 has no way to detect that he is attacking an empty HLL and cannot adapt his strategy, so all buckets hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S4 is able to detect that all buckets are empty and adapt his attack accordingly. The adversary can target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality estimate trade-offs are of course possible.

@@

We challenge the adversary to pick the largest possible number of items to add to the sketch while trying to keep the cardinality estimate as low as possible, either in the S2 or the S4 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.

Those empirical results confirm our impact analysis on several points. First, when attacking a sketch already containing some items, we can see that the adversary is able to hide approximately half of the items' set under S2, and around 83\%%( $205$'$000 \approx 250'000 \times 0.82 \approx 250$'$000\times(1-\frac{1}{2\times H_c})$, with $H_c\approx\frac{1'000}{256}=3.9$).
$\approx (1-\frac{1}{2\times H_c})$ with $H_c\approx\frac{1'000}{256}=3.9$, under S4.
Secondly, when attacking an empty sketch, the final cardinality estimate is always exactly 367 $=\alpha2^{n+1}$ with $\alpha = 0.72$ as specified by the Eq. (29) of \cite{hll2} or 1 in respectively S2 and S4.

@@

\begin{table}[h]
\caption{Result of the attacks over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'000 & 0 & 1'005 & 0 \\ \hline
    \textbf{\# Items added} & 3'655'744 & 3'655'740 & 6'090'003 & 14'220 \\ \hline
    \textbf{Final Est. Card.} & 1'011 & 367 & 1'005 & 1 \\ \hline
\end{tabular}
\label{table:tab1}
\end{table}

\begin{table}[h]
\caption{Result of the attacks in the setting of \cite{hllvuln} over 30 iterations}
\begin{tabular}{| m{8.5em} | m{4em} | m{4em} | m{4em} | m{4em} |}
    \hline
    \textbf{Scenario} & \multicolumn{2}{c|}{S2} & \multicolumn{2}{c|}{S4} \\ \hline
    \textbf{Initial data?} & yes & no & yes & no \\ \hline
    \textbf{Original Est. Card.} & 1'004 & 0 & 998 & 0 \\ \hline
    \textbf{\# Items added} & 124'940 & 125'085 & 204'669 & 501 \\ \hline
    \textbf{Final Est. Card.} & 1'058 & 367 & 998 & 1 \\ \hline
\end{tabular}
\label{table:tab2}
\end{table}
